Cleaning and Processing

### 2024-09-06 code/analysis-code/introanalysis-v1.qmd
Final Revisions for Submission
Cleaned up all files

```{r}
#Path to linear data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_linear.rds")
#load data
ML_linear <- readRDS(data_location)

#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_summary.rds")
#load data
ML_summary <- readRDS(data_location)
```

@@

```{r}
# See LBW and NBW with true denominator
LBW_percentage <- ML_summary %>%
  filter(!is.na(ModifiedColor) & !is.na(BirthweightCategory)) %>% # Remove NAs from both variables

group_by(ModifiedColor) %>% # Group by ModifiedColor
# Calculate the percentage for each group
summarise(
  total_count = n(),
  lbw_count = sum(BirthweightCategory == "LBW"),
  lbw_percentage = (lbw_count / total_count) * 100
  )

# Print the result
print(LBW_percentage)

@@

#Print table
table_fit2

@@

I am not trying to explain which general factors are trying to explain birth weight. Rather, I'm trying to understand if maternal skin color is associated with lower infant birth weight during this specific time period in Rio de Janeiro, Brazil. For example, the gestational age probably explains much of the variation in the model, yet this variable was only included in very few observations in the published clinic data from which I created this dataset.

@@

These results imply that Europeans had heavier infants than Brazilians. Since all women of color in this sample were Brazilian, there is a problem with multicollinearity. So I will perform some subsample modeling to understand this issue. We will create a `Brazilian` (yes, no) variable, drop all non-Brazilians from the study, and run the original regression with the covariates.

@@

Here, non-Brazilians had babies on average weighing 100 grams more than Brazilians (p-value 0.001). So I will now run a regression of only Brazilian women using `Afro-Descent` and `Euro-descent` categories.

@@

>>>>

```{r}
# Path to linear data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_linear.rds")

# Load data
ML_linear <- readRDS(data_location)

# Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_summary.rds")

# Load data
ML_summary <- readRDS(data_location)
```

@@

```{r}
# See LBW and NBW with true denominator
LBW_percentage <- ML_summary %>%
  filter(!is.na(ModifiedColor) & !is.na(BirthweightCategory)) %>% # Remove NAs from both variables
group_by(ModifiedColor) %>% # Group by ModifiedColor

  # Calculate the percentage for each group
summarise(
  total_count = n(),
  lbw_count = sum(BirthweightCategory == "LBW"),
  lbw_percentage = (lbw_count / total_count) * 100
  )

# Print result
print(LBW_percentage)

@@

# Print table
table_fit2

@@

I am not trying to pinpoint which general factors are trying to explain birth weight. Rather, I'm trying to understand if maternal skin color is associated with lower infant birth weight during this specific time period in Rio de Janeiro, Brazil. For example, the gestational age probably explains much of the variation in the model, yet this variable was only included in very few observations in the published clinic data from which I created this dataset.

@@

These results imply that Europeans had heavier infants than Brazilians. Since all women of color in this sample were Brazilian, there is a problem with multicollinearity. So I will perform some sub-sample modeling to understand this issue. We will create a `Brazilian` (yes, no) variable, drop all non-Brazilians from the study, and run the original regression with the covariates.

@@

Here, non-Brazilians had babies on average weighing 100 grams more than Brazilians (p-value = 0.001). So I will now run a regression of only Brazilian women using `Afro-Descent` and `Euro-descent` categories.

@@

Here, I have created a plot of the distribution of birth weights at Laranjeiras, following Wilcox's discussion of the residual and predominant birth weight distributions [@wilcox2001].

### 2024-09-06 code/eda-code/eda-v1.qmd
Final Revisions for Submission
Cleaned up all files

Load the data. First we will load the ML_summary set for summary statistics.

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_summary.rds")

#load data
ML_summary <- readRDS(data_location)
```

@@

# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary, file = summarytable_file)
```

We are saving the results to the `results` folder. All final table/figures will have `_final` in the name. This is to make sure we don't overwrite anything we might want to keep. All supplemental table/figures will have a `suppl` in the name.

@@

Histogram plots for the continuous outcomes.

@@

# Save plot to file
figure_file = here("results","figures", "weight_distribution.png")
ggsave(filename = figure_file, plot=weight_plot, bg="white") #background is white
```
Now infant birth length.
```{r}
# Create histogram plot
length_plot <- ML_summary %>% filter(!is.na(Lengthcentimeters)) %>% #filter out NAs
  ggplot(aes(x = Lengthcentimeters)) +
  geom_histogram(binwidth = 1, fill="#69b3a2") +
  geom_vline(xintercept = c(49), linetype = "dashed", color = "red") +  # Add vertical lines that mark average birth length for both sexes
  theme_ipsum() +
  theme(plot.title = element_text(size=15))
plot(length_plot)

@@

For now, let's look at height as function of weight. We can see that length is positively correlated with weight, which is unsurprising.

@@

Before we create a summary table, lets clean up our variables for easier analysis. First, we need to create dummy variables for categories. We will start with the `Color` variable, which is maternal skin color. We will also create dummy variables for the `Status` and `Nationality` variables for ease of interpretation. The original data categorized maternal racial categories as White (*branca*), the reference group here; mixed-race (*parda*); and Black (*preta*). I maintained this categorization. However, I also created a new dummy variable for skin color, combining Black and mixed-Race patients into one category, `Afro-Descent`, and all White patients into a `Euro-Descent` category.

For the `Status` variable, the Supplementary Table 1 will include all of the gravidity or parity categories from the original data. But I will also create a dummy variable for the summary table for both the final paper and the linear analysis: one category for first pregnancy/birth and one for second or more pregnancies/births. With this, I am removing the distinction between parity and gravidity since it is unclear from the original data if the physicians were categorizing the women before the delivery or after. With this, I am assuming that both gravidity and parity refer to previous births, which risks including a history of stillbirths into a presumed history of live births. However, because these data are about women entering a hospital to give birth, rather than to receive prenatal care for a pregnancy, and this assumption can provide more statistical power for a linear regression, I assume this risk.

@@

For the `Nationality` variable, we will combine all European nationalities into one category, "European", all Latin American nationalities into "Latin American", and all Middle Eastern nationalities into "Middle Eastern". I am combining categories because for many of the `European` and `Latin American` categories, there are very few observations. Moreover, I am interested in immigration as a general status rather than the relationship between specific immigrant countries and health status in Brazil. "Brazilian" will remain the same.

@@

# View the updated dataframe
print(head(ML_summary))
```
I will now change the observations from Portuguese to English for the remaining observations to clean up things for the summary table.

@@

#Summarize missing data using dplyr
missing_data_summary <- ML_summary %>% dplyr::summarise_all(list(~sum(is.na(.))))
missing_data_summary

@@

`Gtsummary` is a very easy way to create summary statistics (Table 1). In the note section at the end of this document, I also include the code I worked on to create the tables manually to show my learning process.

@@

For maternal race, we have a sample size of n = 2695. The majority of the sample is White (branca), with 42.45% of the sample. Mixed-race (parda) is 29.2% of the sample, and Black (preta) is 28.32% of the sample. Let's now compare White patients to all patients of color by combining Black and mixed-race categories. I will create a dummy variable for color, called modified_color with two categories: Afro-Descent (comprising parda and preta) and Euro-Descent (comprising branca).

We can see that of the clinic patients, 1551 (57.6%) were of African descent (defined as preta or parda) and 1144 (42.4%) were of European descent (branca).

We can see that of a sample size of n = 2761, 88% of outcomes were natural deliveries.

Now let's take abortion out of this sample size to just look at labor and delivery outcomes.

```{r}
# Calculate frequencies of birth category without abortion
birth_freq2 <- ML_summary %>%
  filter(!is.na(Birth)) %>% # Exclude rows with NA in birth variable
  filter(Birth != "abortion") %>% # Exclude rows with abortion
  count(Birth, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
print(birth_freq2)
```
We can see that of a new sample size of n = 2672, 91% of outcomes were natural deliveries.

Now let's look at maternal outcomes.

Now let's look at some common reproductive and perinatal health statistics: maternal mortality ratio (MMR), stillbirth rate (SBR), and sex ratio at birth (SRaB). The equation is MMR = MD/LB * 10,000. We can also look at stillbirth rate. The equation is SBR = SB/LB * 1000.
```{r}
#MMR Maternal mortality ratio
maternal_deaths <-23
live_births <- 2445
MMR <- (maternal_deaths / live_births) * 10000
MMR

@@

#Sex ratio at birth (live male births/live female births)*100
sex_ratio <- ML_summary %>%
  filter(!Birth %in% c("Abortion") & !FetalOutcome %in% c("Stillbirth")) %>% # Exclude rows with abortion and stillbirth
 filter(!is.na(Sex) & !is.na(Birth) & !is.na(FetalOutcome)) %>%
  group_by(Sex) %>%
  summarise(count = n()) %>%
  spread(key = Sex, value = count) %>%
  mutate(SexRatio = M / F) * 100
sex_ratio
```
The MMR is 94.1% in this specific clinic. The SBR is 82.9% in this specific clinic. The sex ratio is 1.2 or there were 120 male live births per 100 female live births at this specific clinic. This is much higher than the current range of between 103 and 107 male births per 100 female births.

Now we will create the summary table for mean birth weights across time and place.

@@

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_linear.rds")
#load data
ML_linear <- readRDS(data_location)
```

@@

#Check pre-structure
str(ML_linear)

@@

For the previous code I worked on in relation to summary stats, see the `edacode-extra.R` file.

>>>>

Load the data. First, I will load the ML_summary set for summary statistics.

```{r}
# Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_summary.rds")

# Load data
ML_summary <- readRDS(data_location)
```

@@

# Save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary, file = summarytable_file)
```

I am the results to the `results` folder. All final table/figures will have `_final` in the name. This is to make sure we don't overwrite anything we might want to keep. All supplemental table/figures will have a `suppl` in the name.

@@

Histogram plots for continuous outcomes.

@@

# Save plot to file
figure_file = here("results","figures", "weight_distribution.png")
ggsave(filename = figure_file, plot=weight_plot, bg="white") #background is white
```
Now infant birth length.

```{r}
# Create histogram plot
length_plot <- ML_summary %>% filter(!is.na(Lengthcentimeters)) %>% #filter out NAs
  ggplot(aes(x = Lengthcentimeters)) +
  geom_histogram(binwidth = 1, fill="#69b3a2") +
  geom_vline(xintercept = c(49), linetype = "dashed", color = "red") +  # Add vertical lines that mark average birth length for both sexes
  theme_ipsum() +
  theme(plot.title = element_text(size=15))
plot(length_plot)

@@

For now, let's look at height as function of weight. Length is positively correlated with weight, which is unsurprising.

@@

Before we create a summary table, I will engage in another round of data cleaning for easier analysis. First, I need to create dummy variables for categories. I will start with the `Color` variable, which is maternal skin color. I will also create dummy variables for the `Status` and `Nationality` variables for ease of interpretation. The original data categorized maternal racial categories as White (*branca*), the reference group here; mixed-race (*parda*); and Black (*preta*). I maintained this categorization. However, I also created a new dummy variable for skin color, combining Black and mixed-race patients into one category, `Afro-Descent`, and all White patients into a `Euro-Descent` category.

For the `Status` variable, the Supplementary Table 1 will include all of the gravidity or parity categories from the original data. But I will also create a dummy variable for the summary table for both the final paper and the linear analysis: one category for first pregnancy/birth and one for second or more pregnancies/births. With this, I am removing the distinction between parity and gravidity since it is unclear from the original data if the physicians were categorizing the women before the delivery or after. Thus, I am assuming that both gravidity and parity refer to previous births, which risks including a history of stillbirths into a presumed history of live births. However, because these data are about women entering a hospital to give birth, rather than to receive prenatal care for a pregnancy, and this assumption can provide more statistical power for a linear regression, I assume this risk.

@@

For the `Nationality` variable, I combine all European nationalities into one category, "European", all Latin American nationalities into "Latin American", and all Middle Eastern nationalities into "Middle Eastern". I am combining categories because for many of the `European` and `Latin American` categories, there were very few observations. Moreover, I am interested in immigration as a general status rather than the relationship between specific immigrant countries and health status in Brazil. "Brazilian" will remain the same.

@@

# View the updated dataframe
print(head(ML_summary))
```
I will now change the observations from Portuguese to English for the remaining observations to clean up things for the summary table. In the future, I should perform this step in the processing script.

@@

# Summarize missing data using dplyr
missing_data_summary <- ML_summary %>% dplyr::summarise_all(list(~sum(is.na(.))))
missing_data_summary

@@

`Gtsummary` is a very easy way to create summary statistics (Table 1). For previous code I worked on in relation to summary stats, see the `edacode-extra.R` file.

@@

For maternal skin color, we have a sample size of n = 2695. The majority of the sample is White (*branca*), with 42.45% of the sample. Mixed-race (*parda*) is 29.2% of the sample, and Black (*preta*) is 28.32% of the sample. I'll now compare White patients to all patients of color by combining Black and mixed-race categories. I will create a dummy variable for color, called `ModifiedColor` with two categories: Afro-Descent (comprising *parda* and *preta*) and Euro-Descent (comprising *branca*).

Of the clinic patients, 1551 (57.6%) were of African descent (defined as *preta* or *parda*) and 1144 (42.4%) were of European descent (branca).

Of a sample size of n = 2761, 88% of outcomes were natural deliveries.

I'll now remove abortion from the dataset to look solely at labor and delivery outcomes.

```{r}
# Calculate frequencies of birth category without abortion
birth_freq2 <- ML_summary %>%
  filter(!is.na(Birth)) %>% # Exclude rows with NA in birth variable
  filter(Birth != "abortion") %>% # Exclude rows with abortion
  count(Birth, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
print(birth_freq2)
```

Of the new sample of n = 2672, 91% of outcomes were natural deliveries.

Maternal outcomes are also important to understand, including the following reproductive and perinatal health statistics: maternal mortality ratio (MMR), stillbirth rate (SBR), and sex ratio at birth (SRaB). For MMR, the equation is MMR = MD/LB * 10,000. For SBR, the equation is SBR = SB/LB * 1000. For SRaB, the equation is SRaB = (M/F) * 100.

```{r}
#MMR Maternal mortality ratio
maternal_deaths <-23
live_births <- 2445
MMR <- (maternal_deaths / live_births) * 10000
MMR

@@

#Sex ratio at birth (live male births/live female births)*100
sex_ratio <- ML_summary %>%
  filter(!Birth %in% c("Abortion") & !FetalOutcome %in% c("Stillbirth")) %>% # Exclude rows with abortion and stillbirth
 filter(!is.na(Sex) & !is.na(Birth) & !is.na(FetalOutcome)) %>%
  group_by(Sex) %>%
  summarise(count = n()) %>%
  spread(key = Sex, value = count) %>%
  mutate(SexRatio = M / F) * 100
sex_ratio
```

The MMR is 94.1% in this specific hospital. The SBR is 82.9% in this specific hospital. The sex ratio is 1.2 or there were 120 male live births per 100 female live births at this hospital. This is much higher than the current range of between 103 and 107 male births per 100 female births.

Now I will create the summary table for mean birth weights across time and place.

@@

```{r}
# Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_linear.rds")

# Load data
ML_linear <- readRDS(data_location)
```

@@

```{r}
# Create dummy variables
# Check pre-structure
str(ML_linear)

@@

### 2024-09-06 code/processing-code/processingfile-v1.qmd
Final Revisions for Submission
Cleaned up all files

Please note I used ChatGPT and CoPilot throughout to help troubleshoot code.

# Processing script

This contains the same code and comments/information as `processingcode.R`. This quarto file does not pull code from the R script but rather includes it directly.

@@

```{r}
# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","MaternidadeLaranjeiras.csv")
ML_raw <- read_csv(data_location)

@@

From this initial read, we can see that the data have 2845 observations and 23 variables. The variables are characterized as character and numerical.

@@

First we can look at the Codebook and practice table making.
```{r}
# Path to codebook
codebook_location <- here::here("data", "raw-data", "Codebook.csv")
codebook <- read_csv(codebook_location)

@@

#Print table
print(codebook_table)
```

@@

I created this dataset from published clinical notes from a specific maternity hospital in Rio de Janeiro, Brazil for multiple projects, mostly of a qualitative nature. Thus, I was interested in the notes clinicians included on difficult deliveries and their outcomes. These variables include qualitative information that can be excluded from our processed data, since we are not looking at qualitative notes on cases. Nonetheless, for `Fetal_Notes`, we need to keep the Fetal_Notes variable at the moment as it qualifies the `FetalOutcome` variable.

We also don't need the citation information to run regressions. So we will be removing the following variables as we process our data: `VolN`, `Page`, `Number`, `Nationality_Notes`, `Maternal_Notes`, `CauseofDeath`, `PreviousHistory`, and `Notes`. Further, we will exclude `Birth_Notes`, but first I will need to extract information on gestation (twin versus singleton birth).

@@

Despite removing variables we won't use because of their qualitative nature, we still have many missing observations for some of the variables. Let's check for missing data.

```{r}
#Summarize missing data using dplyr
missing_data_summary <- ML_1 %>% dplyr::summarise_all(list(~sum(is.na(.))))
missing_data_summary
```

We can see that `CivilStatus` has 2834 missing observations, `GestationalAge_Months` has 2789 missing observations, `Fetal_Notes` has 2823 missing observations, and `Birth_Notes` also has 2330 missing values. This is a lot. Although civil status can serve as a proxy for income and class, I have too many missing values here for the variable to be meaningful, so I will exclude `CivilStatus`. Gestational age in months is also a valuable insight in relation to LBW and PTB, but my data set has too many missing values to be of use here, we can also exclude `GestationalAge_Months`. Some gestational age entries were also in relation to spontaneous abortions. `Birth_Notes` helped me initially decided on how to code the `Birth` variable, which was done when I engaged in manual data entry, so I will exclude here.

Before I exclude any variables, I need to make sure that the original hospital records did not misrecord stillbirths as neonatal deaths and vice versa before I exclude them. First, we will look at the values for `FetalOutcome` and `Fetal_Notes` that are not missing any values for both variables. This gives us 22 entries. Most values in the `Fetal_Notes` variable refer to either a macerated (*macerado*) stillbirth, perhaps implying maternal syphilis or a neonatal death (*faleceu*) with a given time frame. We can see that the `FetalOutcome` variable is not always consistent with the `Fetal_Notes` variable. For example, a live birth and subsequent neonatal death is marked as a stillbirth in the `FetalOutcome` variable. I will mark all live births as live, even if death occurred immediately after birth and I will mark all stillbirths as stillbirths. There were 20 recorded neonatal deaths.

@@

I also need to extract information on gestation (twin versus singleton birth) from this `GestationalAge_Months` before I exclude it. My data set will only include singleton births for the statistical analysis.

@@

Now we can check for missing data again.

```{r}
#Summarize missing data using dplyr
missing_data_summary <- ML_4 %>% dplyr::summarise_all(list(~sum(is.na(.))))
missing_data_summary
```
Now, I will create a new categorical variable of normal birth weight (NBW) for all birth weights $>2500$ g and low birth weight (LBW) for all birth weights $\leq2500$ g for a future logistic regression. This will come from the `Weightgrams` variable.

@@

There are still missing observations for all the remaining variables except `Date`. For now, I will keep these observations intact because I want the maximum number of observations for summary statistics of all variables. Although the sample size for these summary statistics will then differ from that of my linear regression model, I believe more observations at the exploratory stage provide a fuller picture of what type of patient went to the hospital during the time period under study. Here, I will create two datasets: one for summary statistics and one for the linear regression model. In addition, most contemporary studies exclude all birthweights less than 500 grams because they have a high chance of being stillborn. I do so the same here, even if they were marked as a live birth. I will also exclude all stillbirths or neonatal deaths as the analysis is interested in studying live births. Finally, for the linear dataset I will exclude all twin births, since most current studies restrict their regression analyses to singleton births.

@@

Now, I can see that my `ML_linear` dataset has no missing data. It has 1919 observations and 12 variables. This might change as I begin analysis. For example, I might want to exclude infant birth length, which is included at the moment.

@@

# Data loading for mean birth weights table
```{r}
# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","Mean_Birthweights.csv")
Mean_Birthweights <- read_csv(data_location)

@@

# Save Mean_Birthweight as RDS
save_data_location <- here::here("data","processed-data","Mean_Birthweights_processed.rds")
saveRDS(Mean_Birthweights, file = save_data_location)
```
# Notes

>>>>

Please note I used ChatGPT, CoPilot, and Claude Anthropic throughout to help troubleshoot code. LLM were *not* used in analysis or writing.

@@

```{r}
# Path to data
# Note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","MaternidadeLaranjeiras.csv")
ML_raw <- read_csv(data_location)

@@

This initial read shows that the data have 2845 observations and 23 variables. The variables are characterized as character and numerical.

@@

First, I will look at the Codebook and practice table making.

```{r}
# Path to codebook
codebook_location <- here::here("data", "raw-data", "Codebook.csv")
codebook <- read_csv(codebook_location)

@@

# Print table
print(codebook_table)
```

@@

I created this dataset from published clinical notes from a specific maternity hospital in Rio de Janeiro, Brazil for multiple projects, mostly of a qualitative nature. Thus, I was interested in the notes clinicians included on difficult deliveries and their outcomes. These variables include qualitative information that can be excluded from our processed data, since we are not looking at qualitative notes on cases. Thus, I will remove the following variables as I process the data: `Nationality_Notes`, `Maternal_Notes`, `CauseofDeath`, `PreviousHistory`, and `Notes`. Nonetheless, for `Fetal_Notes`, we need to keep the Fetal_Notes variable at the moment as it qualifies the `FetalOutcome` variable. Further, I will exclude `Birth_Notes`, but first I will need to extract information on gestation (twin versus singleton birth).

I also don't need the citation information to run regressions. So I will remove the following variables as I process our data: `VolN`, `Page`, `Number`.

@@

Despite removing variables I won't use because of their qualitative nature, I still have many missing observations for some of the variables.

```{r}
# Check for missing data
# Summarize missing data using dplyr
missing_data_summary <- ML_1 %>% dplyr::summarise_all(list(~sum(is.na(.))))
missing_data_summary
```

The `CivilStatus` variable has 2834 missing observations, `GestationalAge_Months` has 2789 missing observations, `Fetal_Notes` has 2823 missing observations, and `Birth_Notes` also has 2330 missing values. Although civil status can serve as a proxy for income and class, I have too many missing values here for the variable to be meaningful, so I will exclude `CivilStatus`. Gestational age in months is also a valuable insight in relation to LBW and PTB, but my data set also has too many missing values, and I will also exclude `GestationalAge_Months`. Some gestational age entries were also in relation to spontaneous abortions. `Birth_Notes` helped me initially decided on how to code the `Birth` variable, which was done when I engaged in manual data entry, so I will exclude here.

Before I exclude any variables, I need to make sure that the original hospital records did not misrecord stillbirths as neonatal deaths and vice versa. First, I will look at the values for `FetalOutcome` and `Fetal_Notes` that are not missing any values for both variables. This gives us 22 entries. Most values in the `Fetal_Notes` variable refer to either a macerated (*macerado*) stillbirth, perhaps implying maternal syphilis, or a neonatal death (*faleceu*) with a specific time frame. Nonetheless, the `FetalOutcome` variable is not always consistent with the `Fetal_Notes` variable. For example, a live birth and subsequent neonatal death is marked as a stillbirth in the `FetalOutcome` variable. I will mark all live births as live, even if death occurred immediately after birth and I will mark all stillbirths as stillbirths. There were 20 recorded neonatal deaths.

@@

I also need to extract information on gestation (twin versus singleton birth) from `GestationalAge_Months` before I exclude it. My data set will only include singleton births for the statistical analysis.

@@

```{r}
# Re-check for missing data
# Summarize missing data using dplyr
missing_data_summary <- ML_4 %>% dplyr::summarise_all(list(~sum(is.na(.))))
missing_data_summary
```
Now, I will create a new categorical variable of normal birth weight (NBW) for all birth weights $>2500$ g and low birth weight (LBW) for all birth weights $\leq2500$ g for a future logistic regression. This will come from the `Weightgrams` variable.

@@

There are still missing observations for all the remaining variables except `Date`. For now, I will keep these observations intact because I want the maximum number of observations for summary statistics of all variables. Although the sample size for these summary statistics will then differ from that of my linear regression model, I believe more observations at the exploratory stage provide a fuller picture of what type of patient went to the hospital during the time period under study. Thus, I will create two datasets: one for summary statistics and one for the linear regression model. In addition, most contemporary studies exclude all birthweights less than 500 grams because they have a high chance of being stillborn. I do so the same here, even if they were marked as a live birth. I will also exclude all stillbirths or neonatal deaths as the analysis is interested in studying live births. Finally, for the linear dataset I will exclude all twin births, since most current studies restrict their regression analyses to singleton births.

@@

Now, I can see that my `ML_linear` dataset has no missing data. It has 1919 observations and 12 variables. This might change as I begin analysis.

@@

# Data loading for mean birth weights table

I am also including a descriptive table of mean birth weights culled from other published papers. Here, I will create a table for these data.

```{r}
# Path to data
# Note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","Mean_Birthweights.csv")
Mean_Birthweights <- read_csv(data_location)

@@

# Save Mean_Birthweight as RDS
save_data_location <- here::here("data","processed-data","Mean_Birthweights.rds")
saveRDS(Mean_Birthweights, file = save_data_location)
```
# Notes
